---
title: "Smoothing and regression splines: Bikes in Washington"
author: "Àlex Martorell, Enric Reverter, Pim Schoolkate"
date: "11/9/2022"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r cars}
rm(list = ls())
load("bikes.Washington.Rdata")
``` 

```{r}
head(bikes)
```

## Question 1

```{r}
sm.spline = smooth.spline(x=bikes$instant, y=bikes$cnt, cv=FALSE)
```

### a) Which is the value of the chosen penalty parameter $\lambda$

```{r}
sm.spline$lambda
```


### b) Which is the corresponding equivalent number of degrees of freedom df?

```{r}
sm.spline$df
```

### c) How many knots have been used

```{r}
sm.spline$fit$nk
```

### d) Give a graphic with the scatter plot and the estimated regression function $\hat{m}$(instant)

```{r}
plot(x=bikes$instant, y=bikes$cnt)
lines(sm.spline$x, sm.spline$y)
```


## Question 2

### a) Use the function logistic.IRWLS.splines to fit the non-parametric binary regression cnt.5000 as a function of the temperature, using df=6.

```{r}
source("IRWLS_logistic_regression.R")
require(dplyr)

bikes = bikes %>%
  mutate(cnt.5000 = case_when(cnt >= 5000 ~ 1,
                              TRUE ~ 0))
sm.spline.binary = logistic.IRWLS.splines(x=bikes$temp, y=bikes$cnt.5000, df=6)
temp.5000 = bikes$temp[which(sm.spline.binary$predicted.values > 0.5)]
```

### a.2) In which range of temperatures is Pr(cnt>=5000|temp) larger than 0,5?

```{r}
min(temp.5000)
max(temp.5000)
```

### b) Choose the parameter df by k-fold log-likelihood cross validation with $k=5$ using df.v = 3:15 as the set of possible values for df.

```{r}
kfold_cv_llk = function(x, y, nfolds, df_vals) {
  
  index = sample(seq_len(nfolds), length(x), replace = TRUE)
  stopifnot(length(x) == length(y))

  llk = numeric(length(df_vals))
  llk_folds = numeric(nfolds)
  i = 1
  
  for (dfi in df_vals) {
    for (n in seq_len(nfolds)) {
      x_training = x[which(index != n)]
      x_testing = x[which(index == n)]
      y_training = y[which(index != n)]
      y_testing = y[which(index == n)]
  
      sm.spline.bin = logistic.IRWLS.splines(x=x_training, y=y_training, x.new=x_testing, df=dfi)
      llk_folds[[n]] = log_likelihood(y_testing, sm.spline.bin$predicted.values)
    }
  llk[[i]] = mean(llk_folds)
  i = i + 1
  }
  return(llk)
}

log_likelihood = function(real, prob) {
  llk = 0
  for (i in 1:length(real)) {
    llk = llk + real[i]*log(prob[i]) + (1-real[i]*log(1-prob[i]))
  }
  return(llk)
}

llk = kfold_cv_llk(x=bikes$temp, y=bikes$cnt.5000, nfolds=5, df_vals=3:15)
```

```{r}
plot(3:15, llk)
lines(3:15, llk)
points(x=which.max(llk)+2, y=max(llk), col="red", cex=2, pch=19)
abline(v=which.max(llk)+2, col="red")
text(x=which.max(llk)+3, y=max(llk), labels=format(max(llk), digits=4), col="red")
```

```{r}
sorted_bikes = bikes %>% arrange(temp)
sm.spline.binary.best = logistic.IRWLS.splines(x=sorted_bikes$temp, y=sorted_bikes$cnt.5000, df=9, plts=TRUE)
```

